\documentname: Is machine learning good or bad for the natural sciences?

### 2024-05-28 paper/main.tex
camera ready, maybe

% This document is Copyright 2024 the author(s).
% It is part of the BadForScience project https://github.com/davidwhogg/BadForScience .

% To do:
% ----------------------------
% - Find the right place to cite the units paper.
% - audit bibliography; make sure arXiv references in the reference list are consistent? Also capitalization!
% - Update appendix with feedback from Vedant Chandra (email). In particular, modify model complexity and information content.
% - Should there be another BAD thing, which is about statistics for comparison of simulations and data? This is a great one; consider after ICML submission.
% - Publish on the arXiv.

@@

\printAffiliationsAndNotice{}
\begin{abstract}
  Machine learning (ML) methods are having a huge impact across all of the sciences.
  However, ML has a strong ontology---in which only the data exist---and a strong epistemology---in which a model is considered good if it performs well on held-out training data.
  These philosophies are in strong conflict with both standard practices and key philosophies in the natural sciences.
  Here we identify some locations for ML in the natural sciences at which the ontology and epistemology are valuable.
  For example, when an expressive machine-learning model is used in a causal inference to represent the effects of confounders, such as foregrounds, backgrounds, or instrument calibration parameters, the model capacity and loose philosophy of ML can make the results more trustworthy.
  We also show that there are contexts in which the introduction of ML introduces strong, unwanted statistical biases.
  For one, when ML models are used to emulate physical (or first-principles) simulations, they introduce strong confirmation biases.
  For another, when expressive regressions are used to label datasets, those labels cannot be used in downstream joint or ensemble analyses without taking on uncontrolled biases.
  The question in the title is being asked of all of the natural sciences; that is, we are calling on the scientific communities to take a step back and consider the role and value of ML in their fields; the (partial) answers we give here come from the particular perspective of physics.
\end{abstract}

@@

As broad as this definition is, it does not include standard model fitting or parameter estimation.
These tasks are usually performed in the context of a well-specified model, where the model does not qualitatively change as more data arrive, and the quantitative improvements to the parameter estimates only scale as something like the square root of the size of the data set.
Standard model fitting and parameter estimation are not ML in our sense.

@@

@@

Finally, we make a philosophical comment:
In the natural sciences, the same phenomena can usually be explained with many qualitatively different theories; theories are not usually unique (eg, \citealt{hogg, peebles}).
In the same way (as we noted above), ML models contain massive degeneracies, and even different architectures can lead to near-identical predictions after training on the same training data.
Does this lead to a philosophical connection between ML and the natural sciences?
We have argued elsewhere (op.cit) that the degeneracies in the natural sciences undermine \emph{realism} or the belief in the literal truth of the latent theories or objects of physics.
There is an argument to be made that maybe the natural sciences could be better off if they moved at least slightly closer to the ML points of view on ontology and epistemology.
In the long run, the data are more stable than the fundamental theories of physics:
Newtonian gravity was replaced with general relativity, and we expect general relativity to be replaced by some kind of quantum gravitational theory.
At each change, the ontology changes, while the data are stable.
We are not recommending anything (yet), we are merely noting that a robust philosophy of natural science might be more positively oriented towards the data than current standard practice.

>>>>

% This document is Copyright 2024 the authors.
% It is part of the BadForScience project https://github.com/davidwhogg/BadForScience .

% To do:
% HOGG: address the comment that standard regression is not ML. (emphasized in yellow comment)
% - Publish on the arXiv.

@@

\printAffiliationsAndNotice{}
\begin{abstract}
  Machine learning (ML) methods are having a huge impact across all of the sciences.
  However, ML has a strong ontology---in which only the data exist---and a strong epistemology---in which a model is considered good if it performs well on held-out training data.
  These philosophies are in strong conflict with both standard practices and key philosophies in the natural sciences.
  Here, we identify some locations for ML in the natural sciences at which the ontology and epistemology are valuable.
  For example, when an expressive machine learning model is used in a causal inference to represent the effects of confounders, such as foregrounds, backgrounds, or instrument calibration parameters, the model capacity and loose philosophy of ML can make the results more trustworthy.
  We also show that there are contexts in which the introduction of ML introduces strong, unwanted statistical biases.
  For one, when ML models are used to emulate physical (or first-principles) simulations, they introduce strong confirmation biases.
  For another, when expressive regressions are used to label datasets, those labels cannot be used in downstream joint or ensemble analyses without taking on uncontrolled biases.
  The question in the title is being asked of all of the natural sciences; that is, we are calling on the scientific communities to take a step back and consider the role and value of ML in their fields; the (partial) answers we give here come from the particular perspective of physics.
\end{abstract}

@@

As broad as this definition is, it does not include standard model fitting or parameter estimation with a well-specified (first-principles, say) model.
These tasks are usually performed in the context of a rigid model, with far fewer parameters than data points, in which the model does not qualitatively change as more data arrive, and the quantitative improvements to the parameter estimates only scale as something like the square root of the size of the data set.
Standard model fitting and parameter estimation in the enormously under-parameterized regime are not ML in our sense.

@@

A recent position piece \citep{donoho2024data} points out that an important aspect of ML's epistemology might be its (aspirational) \emph{reproducibility}.
Namely, many contributions to the field permit anyone to reproduce the results and build on top of them; the idea is that
this aspect of ML is part of the reason that the field moves so fast.
In contrast, it is often argued that the sciences face a reproducibility crisis \cite{baker2016reproducibility};
the (only) positive spin to put on that is that reproducibility in the natural sciences is not nearly as straightforward as it is in ML.
This reproducibility discussion is slightly out-of-scope here; we refer the reader to Donoho's piece and the follow-up commentaries (for example, \citealt{Freire2024Singularity}, \citealt{Milanfar2024Data}).

@@

Finally, we make a philosophical comment:
In the natural sciences, the same phenomena can usually be explained with many qualitatively different theories; theories are not usually unique (eg, \citealt{hogg, peebles}).
In the same way (as we noted above), ML models contain massive degeneracies, and even different architectures can lead to near-identical predictions after training on the same training data.
Does this lead to a philosophical connection between ML and the natural sciences?
We have argued previously \cite{hogg} that the degeneracies in the natural sciences undermine \emph{realism} or the belief in the literal truth of the latent theories or objects of physics.
There is an argument to be made that maybe the natural sciences could be better off if they moved at least slightly closer to the ML points of view on ontology and epistemology.
In the long run, the data are more stable than the fundamental theories of physics:
Newtonian gravity was replaced with general relativity, and we expect general relativity to be replaced by some kind of quantum gravitational theory.
At each change, the ontology changes, while the data are stable.
We are not recommending anything (yet), we are merely noting that a robust philosophy of natural science might be more positively oriented towards the data than current standard practice.
